Scaling up a GF grammar for free parsing

### 2011-12-11 doc/report/report.tex
related work section

\tableofcontents
\newpage
\chapter{Introduction}
>> short intro to GF
The grammar formalism GF has so far been succesfull used for 
controlled languges, but recent experiments have showed
that it is also possible to use the framework for parsing free language.
Parsing, that is the task of automatically identifiyng morpohlogical and
syntactical structures to text, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that give a deep analyse for
Swedish. The fact that the existing rule-based Swedish
parsers are not freely available does not only restrict their
usage but also their possiblities of being developed. Our
goal is
to create a open source Swedish grammar and parser that can be continuosly enhanced and
be made use of in other projects.\\
To build a parser or a grammar from scratch is a heavy work ~~ and we
therefore combine already existing parts of ??.
We start from GF resource grammar and thereby get
%The usage of GF allows us to start from
a well-defined system for describing grammar, as well as a strong connection to
more than 20 other langugages implentented in the framework.
>>The grammar defines a parser, which accepts all constructions defined by the grammar.

\section{Aims}
>> enhance gf...
The purpose of this project has been to prepare the GF grammar for free parsing.
This has meant to develop earlier techniques to fit for Swedish, create methods
for achieving and keeping a extensive lexicon and to adapt the existing
resource grammar to fit for more language specific constrictions.
We hence needed to extend the Swedish grammar to cover
the language specific constructions. Further a large lexicon was
needed, as well as a method for evaluation.
During this project, we have worked on and explored those things and
while identifying problems and future work.

@@

\section{Grammatical Framework}
\label{sec:gf}
The main component of the the project is the Grammatical Framework (GF) \cite{gfbok}, %It is
a grammar formalism based on Martin-Löf type theory \cite{martinlof}. GF is
designed for multilingual applications and the formalism is equivalent to
Parallel Multiple Context-Free Grammars (Seki \& al. 1991), it is hence
stronger than mildly contex-free grammars and can express whatever tree
adjoining grammars (Joshni,xx) and HG?? can.  --very confusing here!!

GF is a strongly typed functional programming languages, inspired by
%programming languages like 
ML \cite{ml} and Haskell \cite{haskell}. Its type
checking system is inspired by \textlambda Prolog \cite{bratko} and the
dependent types by Agda \cite{agda}. 
>> check this!!
The first version of the framework was implemented in 19xx in xx and has
since then fatt en trogen skara användare, med headquarter 
at Gothenburg University.

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. >>check!
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF)\ref{pgf},
which can be used with %There are libraries for using PGF files with
Java, Java script, C and Haskell. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to try and test her grammar by using commands for parsing, vizualization of parse trees,
random generation, word alignment, morphological quizzes etc.
%The following example shows how to parse a sentence and visualizing the parse tree.
%, by using the \textit{pipe} \verb-|-, sending the result to 
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_parse
> parse "jag ser katten" | visualize_tree
\end{verbatim}
\caption{Example of how the GF shell can be used. The first command will parse the given
sentence and visualize the resulting parse tree and the second will visualize the 
abstract syntax tree.}
\label{fig:shellvp}
\end{figure}
>> test vp -format=pdf -view=evince
>> test vt -format=pdf -view=evince -nofun

The visualization of the trees is done with graphviz\footnote{http://www.graphviz.org/}.
Figure \ref{fig:gftree1} shows the results from running the commands in figure \ref{fig:shellvp}
A GF parse tree has two visualized representations, the \textit{parse tree} and
the \textit{abstract tree}.
\begin{figure}[h]
\centering
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=60mm]{gfTree.pdf}}
\subfloat[GF parse tree]{\label{pic:gfStree}\includegraphics[width=50mm]{gfPTree.pdf}}
\subfloat[GF parse tree]{\label{pic:gfEtree}\includegraphics[width=50mm]{gfETree.pdf}}
\caption{Parse tree and abstract tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}
The abstract trees shows all functions GF needs to use
to parse the sentence.
% do not know about abstract/concrete yet: and is consequently the same for all concrete syntaxes.
The parse tree (\ref{pic:gfStree}) shows the words and which types the functions assign
to them.
The reader should keep in mind that the visualized parse trees is not a
complete representation.
If the end node of a  branch has no corresponding word, it is excluded from
the visualized parse tree. This is the case for the definite article in Swedish,
and the information \verb-DetQuant DefArt NumSg- is not shown in figure \ref{pic:gfStree}.
In the corresponding English parse, figure \ref{pic:gfEtree},
the article \emph{``the"} is explicit and therefore shown in the parse tree.

@@

%>>the scope is controlled language, but possible ambiguities (discuss this in future
%>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
%>>syntactical (man with the telescope), lexical (youSg/Pl, växa).
%>> have ambiguity section. 
The scope of GF grammars is usually controlled language. By
restricting the coverage, the number of ambiguities
can be limited and we can ensure that the semantics is preserved during
translation. The ambiguities arising from multilinguality and the inherited will of
course remain, but limited. %Example of domainspecific word?
The use of controlled language gives a possibility of good and reliable translation.
Some examples of projects using GF are WebAlt \cite{webalt},
a project which aims to develop language independent material for mathematics problems,
and the formal software verification tool KeY \cite{key}, in wich GF is used for
translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
The framework is also used in the
European project MOLTO \cite{molto} for online translation between up to 15 language.\\

\subsection{Writing a GF grammar}
The key feature in GF is the distinction between
\textit{abstract} and \textit{concrete} syntax. The abstract syntax represents
the internal structure and models the semantics without concern for language
specific features such as agreement or word order.
One abstract grammar can be implemented by a number of concrete grammars, each
representing one language. As a comparison, the abstract and concrete syntax
may be thought of as fstructures and cstructures in LFG (x,x).

@@

\subsection{The resource library}
%>>Best part? Biggest library?
%>>Most general resource, first large grammar, covers fundamental syntactics.
The GF package provides a powerful resource library \cite{gf-resource}, which covers the
fundamental morphology and syntactics for more than 20 languages. %is a very useful part of the framework. It is provided in the 
%, they give a good start for all who want to write an application grammar.
%of a language for each application, all
%this can be found as libraries. So far there are 20 languages in the library, and
The languages all share the same abstract and they constitute  
a very useful resource when implementing application grammars.
There is also a small test lexicon included, containing some hundred of common
words.

The resource grammars describe how to construct phrases and sentences and how to
decline words. The latter is done by smart paradigms: functions that analyse
some given forms of a word to find out how the inflection table should look.
For example, the declination of many Swedish nouns can be determined by looking
at the singular form only. 
\begin{wrapfigure}{l}{0.4\textwidth}
\begin{tabular}{| l | l |}
\hline
1st declination & 5th declination \\
\hline
flick\textbf{a}    &     hjärt\textbf{a}   \\
flick\textbf{an}    &    hjärt\textbf{at}  \\
flick\textbf{or}    &    hjärt\textbf{an}  \\
flick\textbf{orna}  &    hjärt\textbf{ana} \\
\hline
\end{tabular}
\end{wrapfigure}
This is the case for ``flicka", the correct declination
can be guessed by using the first declination. % and noting that the word ends with an ``a". 
%Like most Swedish nouns ending with an ``a", it follows the first declination.
For others, like "hjärta", also the plural form "hjärtan" is needed.
%, even though it ends with an ``a". 
The worst case is nouns that need four forms, both singular and
plural in definite and indefinite form.

Section \ref{sec:swegf} will give a more thorough description of the Swedish resource grammar.

@@

\section{Related work}
\label{sec:related}
>> Much more here, important!! comparisons.
>> HPSG and LFG are context sensitive
Computational ling is active area for Swedish, there are many other Swedish parser,
but grammars?
Many of the parsers not ruled based, the most known one Malt \cite{malt} ... .
However has been two, about CLE \cite{cle} and what happened to it. Logic and translation
to Swedish, S-CLE (Gambäck,1997) full analyse and semantics, translation,
but coverage is too limited (nivres papper), The
spoken lang. translator \cite{cle2}. Wiren \cite{wiren}?? chartparser, but for swedish. bit more information than
just pos: cassSwe \cite{casswe}
Other Sw. parsers and grammars, evaluation of them? 
Other grammar frameworks.
SweCG (birn) annotation, parser Swedish FDG (voultanien,2001) which is not freely available.
finding subject, objects, suubject complements (SC) and
to link them to their proper regents (main verbs) : over 90\% both precision and recall.
uses a dependency grammar, but "The functional description of adverb phrases
and prepositional phrases (e.g. agent, source, goal, benefactive, time) remains
to be described in a future version." 
HPSG \cite{hpsg} and lingo matrix \cite{lingomatrix}, English, Japanese, German,
Greece, Spanish, Norweigan (NorSource, Hellan 2003),( not famous: SweCore
(Ahrenberg) based on NorSource)
also Swedish: BiTSE \cite{stymne}, Stymne 2006, Swedish English translation,
and one for the common parts of scandinavian (Haugereid etc) in Nordisk sprog.
focusing on verb phrames, built on HPSG.
S\o rgaard and Haugereid, Scandinavian Grammar Matrix \cite{scandmatrix}
Tag, Tree adjoining grammar, XTAG: English and Korean, Chinese? English: ongoing to get widecoverage grammar
under GPL. Translation to Korean.
% related and references work: CLE, Krasimirs book, CassSwe, GF book,SAG, gunlög,
% the multilingual treebank, Homse&Hinchcliff, Aarne i LILT

>>>>

\tableofcontents
\newpage
\chapter{Introduction}
%>> short intro to GF
The grammar formalism GF has so far been succesfull used for 
controlled languges, but recent experiments have showed
that it is also possible to use the framework for parsing free language.
Parsing, that is the task of automatically identifiyng morpohlogical and
syntactical structures to text, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that give a deep analyse for
Swedish. The fact that the existing rule-based Swedish
parsers are not freely available does not only restrict their
usage but also their possiblities of being developed. Our
goal is
to create a open source Swedish grammar and parser that can be continuosly enhanced and
be made use of in other projects.\\
To build a parser or a grammar from scratch is a heavy work ~~ and we
therefore combine already existing resources.
We start from GF resource grammar and thereby get
%The usage of GF allows us to start from
a well-defined system for describing grammar, as well as a strong connection to
more than 20 other langugages implentented in the framework.
%>>The grammar defines a parser, which accepts all constructions defined by the grammar.

\section{Aims}
%>> enhance gf...
The purpose of this project has been to prepare the GF grammar for free parsing.
This has meant to develop earlier techniques to fit for Swedish, create methods
for achieving and keeping a extensive lexicon and to adapt the existing
resource grammar to fit for more language specific constrictions.
We hence needed to extend the Swedish grammar to cover
the language specific constructions. Further a large lexicon was
needed, as well as a method for evaluation.
During this project, we have worked on and explored those things and
while identifying problems and future work.

@@

\section{Grammatical Framework}
\label{sec:gf}
The main component of the the project is the Grammatical Framework (GF) \cite{gfbok}, %It is
a grammar formalism based on Martin-Löf type theory \cite{martinlof}. GF is
designed for multilingual applications and the formalism is 
%equivalent to Parallel Multiple Context-Free Grammars (Seki \& al. 1991), an instance of
%generalized context free grammar (GCFG) it is hence
stronger than mildly contex-free grammars and hence has the same expressiveness
as Tree Adjoining Grammars (Joshi,1975) and Head Grammars (Pollard,1984).

GF is a strongly typed functional programming languages, inspired by
%programming languages like 
ML \cite{ml} and Haskell \cite{haskell}. The logic and reasoning 
are inspired by \textlambda Prolog \cite{prolog} and 
by Agda \cite{agda}, with which GF also shares its type checking algorithm.
The first version of the framework was implemented by the Xerox Research Center
in Grenoble and is now mainly developed in Gothenburg. On of the biggest
achivements is a library covering the 
basic morphological and syntactical structures of more than
20 languages (see section \ref{sec:resources}).

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
which can be used with %There are libraries for using PGF files with
Java, Java script, C and Haskell. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to try and test her grammar by using commands for parsing, vizualization of parse trees,
random generation, word alignment, morphological quizzes etc.
%The following example shows how to parse a sentence and visualizing the parse tree.
%, by using the \textit{pipe} \verb-|-, sending the result to 
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used. The first command will parse the given
sentence and visualize the resulting parse tree and the second will visualize the 
abstract syntax tree.}
\label{fig:shellvp}
\end{figure}
\begin{figure}[h]
\centering
\subfloat[GF abstract tree]{\label{pic:gfAtree}\includegraphics[width=100mm]{gfTree.pdf}}
\subfloat[Swedish parse tree]{\label{pic:gfStree}\includegraphics[width=30mm]{gfSTree.png}}
\caption{Parse tree and abstract tree for the sentence ``Jag ser katten".}
\label{fig:gftree1}
\end{figure}

% nice: test vt -format=pdf -view=evince -nofun
\begin{wrapfigure}{l}{0.3\textwidth}
\includegraphics[width=20mm]{gfETree.png}
\caption{English parse tree}
\label{pic:gfEtree}
\end{wrapfigure}

The visualization of the trees is done with graphviz\footnote{http://www.graphviz.org/}.
Figure \ref{fig:gftree1} shows the results from running the commands in figure \ref{fig:shellvp}
A GF parse tree has two visualized representations, the \textit{parse tree} and
the \textit{abstract tree}.
The abstract trees shows all functions GF needs to use
to parse the sentence.
% do not know about abstract/concrete yet: and is consequently the same for all concrete syntaxes.
The parse tree (\ref{pic:gfStree}) shows the words and which types
the functions assign to them.
The reader should keep in mind that the visualized parse trees is not a
complete representation.
If the end node of a  branch has no corresponding word, it is excluded from
the visualized parse tree. This is the case for the definite article in Swedish,
and the information \verb-DetQuant DefArt NumSg- is not shown in figure \ref{pic:gfStree}.

In the corresponding English parse tree, figure \ref{pic:gfEtree},
the article \emph{``the"} is explicit and therefore shown as a quantifier in the parse tree.\\

@@

%>>the scope is controlled language, but possible ambiguities (discuss this in future
%>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
%>>syntactical (man with the telescope), lexical (youSg/Pl, växa).
%>> have ambiguity section. 
The scope of GF grammars is usually controlled language. By
restricting the coverage, the number of ambiguities
can be limited and we can ensure that the semantics is preserved during
translation. The inhereted ambiguities and the ones arising from multilinguality will
of course remain, but can be controlled more easily. %Example of domainspecific word?
The use of controlled language gives a possibility of good and reliable translation.
  Some examples of projects using GF are WebAlt \cite{webalt},
a project which aims to develop language independent material for mathematics problems,
and the formal software verification tool KeY \cite{key}, in wich GF is used for
translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
The framework is also used in the
European project MOLTO \cite{molto} for online translation between up to 15 language.\\

\subsection{Writing a GF grammar}
The key feature in GF is the distinction between
\textit{abstract} and \textit{concrete} syntax. The abstract syntax represents
the internal structure and models the semantics without concern for language
specific features such as agreement or word order.
One abstract grammar can be implemented by a number of concrete grammars, each
representing one language. As a comparison, the abstract and concrete syntax
may be thought of as fstructures and cstructures in Lexical Functional Grammar (Bresnan and Kaplan, 1982).

@@

\subsection{The resource library}
\label{sec:resources}
%>>Best part? Biggest library?
%>>Most general resource, first large grammar, covers fundamental syntactics.
The GF package provides a powerful resource library \cite{gf-resource}, which covers the
fundamental morphology and syntactics for more than 20 languages. %is a very useful part of the framework. It is provided in the 
%, they give a good start for all who want to write an application grammar.
%of a language for each application, all
%this can be found as libraries. So far there are 20 languages in the library, and
The languages all share the same abstract and they constitute  
a very useful resource when implementing application grammars.
There is also a small test lexicon included, containing some hundred of common
words.

The resource grammars describe how to construct phrases and sentences and how to
decline words. The latter is done by smart paradigms: functions that analyse
some given forms of a word to find out how the inflection table should look.
For example, the declination of many Swedish nouns can be determined by looking
at the singular form only. 
\begin{wrapfigure}{l}{0.4\textwidth}
\begin{tabular}{| l | l |}
\hline
1st declination & 5th declination \\
\hline
flick\textbf{a}    &     hjärt\textbf{a}   \\
flick\textbf{an}    &    hjärt\textbf{at}  \\
flick\textbf{or}    &    hjärt\textbf{an}  \\
flick\textbf{orna}  &    hjärt\textbf{ana} \\
\hline
\end{tabular}
\end{wrapfigure}
This is the case for ``flicka", the correct declination
can be guessed by using the first declination. % and noting that the word ends with an ``a". 
%Like most Swedish nouns ending with an ``a", it follows the first declination.
For others, like "hjärta", also the plural form "hjärtan" is needed.
%, even though it ends with an ``a". 
The worst case is nouns that need four forms, both singular and
plural in definite and indefinite form.
Section \ref{sec:swegf} will give a more thorough description of the Swedish resource grammar.

@@

\section{Related work}
\label{sec:related}
There has been much research about parsing Swedish. One of the most well known
parser is the data-driven Malt parser \cite{malt}, which has been trained on 
Talbanken. 
There are also a number of grammar-based parsers, although none is still freely avaible.
CassSwe \cite{casswe} is a cascaded finite state parser giving a syntactic
analysis.
The Swedish Constraint Grammar 
(Birn,1998) gives a shallow syntactic analysis.
Swedish FDG (Voultanien,2001) uses the Functional Dependency Grammar
(Tapanainen and Järvinen,1997), an extension of the Constraint Gramman
formalism, and produces a dependecy structure focusing on finding the nominal
arguments. \\
%for the  finding subject, objects, suubject complements (SC) and
%to link them to their proper regents (main verbs) : over 90\% both precision and recall.

%Tag, Tree adjoining grammar, XTAG: English and Korean, Chinese? English:
%ongoing to get widecoverage grammar
%under GPL. Translation to Korean.

The Lingo Matrix \cite{lingomatrix}, is a starter-kit for building Head-Driven Phrase
Structure Grammars \cite{hpsg} (HPSG) providing comptability with tools for
parsing, evaluation, semantic representations etc.
There is a collection of implementations done in this framework?, describing ie.
English, Japanese and German. %, Greece, Spanish etc. 
The Scandinavian Grammar Matrix \cite{scandmatrix} covers common parts of
Scandiavian, while Norsource (Hellan, 2003) describes Norweigan. A Swedish version
was based upon this (SweCore, Ahrenberg) which covers the morphology and some
differences between Swedish and Norweigan. Further, there is the BiTSE \cite{stymne}
grammar, also implemented using the Lingo Matrix,
which focuses on verb frames and translation of those by using Minimal Recursion
Semantics \cite{mrs} as an interlingua. \\
It should be pointed out that grammars implemented using Lingo Matrix
do not share any common abstract, like the resources grammars of GF?\\

The Swedish verison \cite{gamback} of the Core Language Enigne (CLE) \cite{cle}
gives a full syntacitc analysis as well as semantics represented in xx. A
translation for conversations?? to English was implemented. The work was further developed in the spoken
language blabla (xx,xx), but is no longer available. The coverage of the Swedish
CLE is also reported to be very limited (ref).\\

In the TAG formalism (Joshi,1975), there are projects on getting open source, widecoverage grammars
for ie. English and Korean, but, to our knowledge, not for Swedish.  
%However has been two, about CLE \cite{cle} and what happened to it. Logic and translation
%to Swedish, S-CLE (Gambäck,1997) full analyse and semantics, translation,
%but coverage is too limited (nivres papper), The
%spoken lang. translator \cite{cle2}.

%>> Much more here, important!! comparisons.
%>> HPSG and LFG are context sensitive
%Computational ling is active area for Swedish, there are many other Swedish parser,
%but grammars?
%Many of the parsers not ruled based, the most known one Malt \cite{malt} ... .
%Wiren \cite{wiren}?? chartparser, but for
%swedish. bit more information than just pos: cassSwe \cite{casswe}
%Other Sw. parsers and grammars, evaluation of them? 
%Other grammar frameworks.
%SweCG (birn) annotation, parser Swedish FDG  (voultanien,2001)which is not
%freely available.
%finding subject, objects, subject complements (SC) and
%to link them to their proper regents (main verbs) : over 90\% both precision and recall.
%uses a dependency grammar, but "The functional description of adverb phrases
%and prepositional phrases (e.g. agent, source, goal, benefactive, time) remains
%to be described in a future version." 
%HPSG \cite{hpsg} and lingo matrix , English, Japanese, German,
%Greece, Spanish, Norweigan (NorSource, Hellan 2003),( not famous: SweCore
%(Ahrenberg) based on NorSource)
%also Swedish: BiTSE \cite{stymne}, Stymne 2006, Swedish English translation,
%and one for the common parts of scandinavian (Haugereid etc) in Nordisk sprog.
%focusing on verb phrames, built on HPSG.
%S\o rgaard and Haugereid, Scandinavian Grammar Matrix \cite{scandmatrix}
% related and references work: CLE, Krasimirs book, CassSwe, GF book,SAG, gunlög,
% the multilingual treebank, Homse&Hinchcliff, Aarne i LILT
